\sysfull

### 2018-04-16 paper/implementation.tex
paging
revisions to implementation

Table~\ref{table:coala_api} summarizes \sys's API methods and their arguments. \todo{Add example of a Coala program}{Carlo, Przemek} These methods trigger specific \sys kernel functions, whose explanation is given in the following sections.

@@

\begin{wrapfigure}{t!}{0.5\textwidth}
	\centering
	%\includegraphics[width=0.5\columnwidth]{figures/graffle/state-machine.pdf}
	\caption{\sys state machine implementation \todo{Draw a figure}{Sinan, Carlo}}
	\label{fig:coala_state_machine}
\end{wrapfigure}

\textbf{Execution.} The application control is passed to \sys's execution manager by calling \texttt{COALA\_RUN} at the end of the main function. A high-level state machine of the execution manager is depicted in Figure~\ref{fig:coala_state_machine} \todo{Draw a figure reflecting the following description}{Sinan, Carlo}. Upon reboot, the coalescing target is initialized according to the coalescing strategy. Then, the second phase of the commit is resumed, in case the last power failure occurred during this operation. The shadow buffer list is cleared, and the metadata of the volatile buffer is initialized. The pointer to the last coalescing task is recalled from non-volatile memory. Before running, \sys checks whether there is any partial execution to resume, in which case the volatile state, including the program counter, is restored. Otherwise, coalescing starts: the current budget is replenished; tasks are run, the history is increased and the current budget is decreased; when the latter is zero, coalescing stops, the pointer to the next task is saved and the commit phase starts. Coalescing resumes after commit phase two ends and the target is updated based on the coalescing strategy. \todo{Go over keywords (coalescing target, current budget, commit phases etc.) and make them consistent with the rest of the paper}{Amjad, Carlo}

@@

When a \sys application accesses a protected variable with either \texttt{RP} or \texttt{WP}, \sys has to search for the variable in RAM first, and then, if not found, in non-volatile memory. In reality, the page manager searches for pages, not single variables. A page search can be very fast if each page is identified by a unique tag, and if every variable belonging to a page bears the tag in its address. Thus, retrieving a page tag from a protected variable's address boils down to extracting the upper bits of the address itself. The remaining lower bits denote the variable's offset in its page. The complete process is depicted in Figure~\ref{figure:coala_page_tags} \todo{this figure doesn't really show page tags and offsets, but paging in general}.
We note that, in order to obtain unique page tags, some alignment requirements have to be met. First, the page size $S$ has to be a power of two. Moreover, pages have to be aligned by $S$. Note that, by aligning pages by $2^n$, the alignment is also valid for all sizes $2^x$, where $x \leq n$. For our experiments, we used 32-, 64-, 128- and 256-byte pages. 16\,KB were allocated in non-volatile memory for \texttt{\underline{shadow}} and \texttt{\underline{store}} (8\,KB each), and the volatile page buffer was 1\,KB, half the total SRAM size. The main advantage of having large pages is searches through page buffers are shorter. On the other hand, page swaps are more bulky.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figures/graffle/paging.pdf}
	\caption{Illustration of \sys page tags mechanism.}
	\label{figure:coala_page_tags}
\end{figure}

@@

\subsection{Partial execution}

Among \sys's novelties, the task-splitting feature is the most effective when it comes to preservation of forward progress. Its implementation adds hardly any computational overhead, and is supported by a hardware timer running "in background", whose energy consumption is extremely low (around $3 \mu A/MHz$~\cite{msp430datasheet}). The goal is to interrupt a static task if it cannot be run to completion with a full capacitor's energy, commit its partial results, save volatile state and program counter and continue. Coala speculates and assumes that a static task is too large and has to be split when power fails two times consecutively in the same static task, and the coalescing target is one, because, after the first power failure, the large task is run with a fresh energy budget. To "split" a static task, Coala schedules a timer interrupt. If this interrupt occurs before the capacitor is depleted, Coala commits all the dirty volatile pages (using the two-phase commit explained above), saves stack and volatile registers to non-volatile memory and then continues with the task. In case a power failure occurs after this operation, but before the static task completes, the execution at the next reboot starts exactly where Coala had saved the partial execution. A task can be split in multiple sub-tasks, as many as needed to progress forward. Regarding the timer interrupt, it is initially scheduled to fire after a large time interval. If power fails before the interrupt fires, Coala reduces the interval. If the partial execution is committed successfully, Coala enlarges the interval. These adjustments follow a binary search technique, and make the interval converge very fast to a value optimal for preserving forward progress and avoiding redundant commits.

@@

\begin{wrapfigure}{t!}{0.5\textwidth}
	\centering
	%\includegraphics[width=0.5\columnwidth]{figures/graffle/commits-illustration.pdf}
	\caption{Intermediate commit and ultimate commit Illustration \todo{to be drawn}{sinan}}
	\label{fig:intermediate_ultimate-commit}
\end{wrapfigure}

>>>>

Table~\ref{table:coala_api} summarizes \sys's API methods and their arguments. These methods trigger specific \sys kernel functions, whose explanation is given in the following sections.

@@

%\begin{wrapfigure}{t!}{0.5\textwidth}
%	\centering
%	%\includegraphics[width=0.5\columnwidth]{figures/graffle/state-machine.pdf}
%	\caption{\sys state machine implementation \todo{Draw a figure}{Sinan, Carlo}}
%	\label{fig:coala_state_machine}
%\end{wrapfigure}

%A high-level state machine of the execution manager is depicted in Figure~\ref{fig:coala_state_machine} \todo{Draw a figure reflecting the following description}{Sinan, Carlo}.

\textbf{Execution.} The application control is passed to \sys's execution manager by calling \texttt{COALA\_RUN} at the end of the main function. Upon reboot, the coalescing target is initialized according to the coalescing strategy. Then, the second phase of the commit is resumed, in case the last power failure occurred during this operation. The shadow buffer list is cleared, and the metadata of the volatile buffer is initialized. The pointer to the last coalescing task is recalled from non-volatile memory. Before running, \sys checks whether there is any partial execution to resume, in which case the volatile state, including the program counter, is restored. Otherwise, coalescing starts: the current budget is replenished; tasks are run, the history is increased and the current budget is decreased; when the latter is zero, coalescing stops, the pointer to the next task is saved and the commit phase starts. Coalescing resumes after commit phase two ends and the target is updated based on the coalescing strategy. \todo{Go over keywords (coalescing target, current budget, commit phases etc.) and make them consistent with the rest of the paper}{Amjad, Carlo}

@@

When a \sys application accesses a protected variable with either \texttt{RP} or \texttt{WP}, \sys has to search for the variable in RAM first, and then, if not found, in non-volatile memory. In reality, the page manager searches for pages, not single variables. A page search can be very fast if each page is identified by a unique tag, and if every variable belonging to a page bears the tag in its address. Thus, retrieving a page tag from a protected variable's address boils down to extracting the upper bits of the address itself. The remaining lower bits denote the variable's offset in its page. 
%The complete process is depicted in Figure~\ref{figure:coala_page_tags} \todo{this figure doesn't really show page tags and offsets, but paging in general}.
We note that, in order to obtain unique page tags, some alignment requirements have to be met. First, the page size $S$ has to be a power of two. Moreover, pages have to be aligned by $S$. Note that, by aligning pages by $2^n$, the alignment is also valid for all sizes $2^x$, where $x \leq n$. For our experiments, we used 32-, 64-, 128- and 256-byte pages. 16\,KB were allocated in non-volatile memory for \texttt{\underline{shadow}} and \texttt{\underline{store}} (8\,KB each), and the volatile page buffer was 1\,KB, half the total SRAM size. The main advantage of having large pages is searches through page buffers are shorter. On the other hand, page swaps are more bulky.

%\begin{figure}
%	\centering
%	\includegraphics[width=\textwidth]{figures/graffle/paging.pdf}
%	\caption{Illustration of \sys page tags mechanism.}
%	\label{figure:coala_page_tags}
%\end{figure}

@@

\subsection{Partial Execution}

Among \sys's novelties, the task-splitting feature is the most effective when it comes to preservation of forward progress. Its implementation adds hardly any computational overhead, and is supported by a hardware timer running ``in the background'', whose energy consumption is extremely low, i,e. $\approx$3\,$\mu$ A/MHz~\cite{msp430datasheet}. The goal is to interrupt a static task if it cannot be run to a completion with a full capacitor's energy, commit its partial results, save volatile state and program counter and continue. \sys speculates and assumes that a static task is too large and has to be split when power fails two times consecutively in the same static task, and the coalescing target is one, because, after the first power failure, the large task is run with a fresh energy budget. To ``split'' a static task, \sys schedules a timer interrupt. If this interrupt occurs before the capacitor is depleted, \sys commits all the dirty volatile pages (using the two-phase commit explained above), saves stack and volatile registers to non-volatile memory and then continues with the task. In case a power failure occurs after this operation, but before the static task completes, the execution at the next reboot starts exactly where \sys had saved the partial execution. A task can be split in multiple sub-tasks, as many as needed to progress forward. Regarding the timer interrupt, it is initially scheduled to fire after a large time interval. If power fails before the interrupt fires, \sys reduces the interval. If the partial execution is committed successfully, \sys enlarges the interval. These adjustments follow a binary search technique, and make the interval converge very fast to a value optimal for preserving forward progress and avoiding redundant commits.

@@

%\begin{wrapfigure}{t!}{0.5\textwidth}
%	\centering
%	%\includegraphics[width=0.5\columnwidth]{figures/graffle/commits-illustration.pdf}
%	\caption{Intermediate commit and ultimate commit Illustration \todo{to be drawn}{sinan}}
%	\label{fig:intermediate_ultimate-commit}
%\end{wrapfigure}
